---
title: "Prosper Loan Exploration"
author: "Andrew J. Gill"
date: "March 31, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(reshape2)
library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(dplyr)

getwd()
setwd("C:/Users/Andrew/Documents/Udacity/Data_Vis")
loan = read.csv("prosperLoanData.csv")
prosper <- loan[, c("BorrowerAPR", "ProsperScore", "LoanOriginalAmount", "Term", 'ListingCreationDate')]
prosper$year <- substring(prosper$ListingCreationDate, 0, 4)
loan$year <- substring(loan$ListingCreationDate, 0, 4)
prosper <- na.omit(prosper)

write.csv(prosper, file="prosper.csv", row.names=F)
write.table(prosper, file="prosper.tsv", quote=F, sep='\t', row.names=F)


apr_by_year_score <- group_by(loan, ProsperScore, year) %>%
                summarise(mean_apr= mean(BorrowerAPR),
                n = n()) %>%
                ungroup()

apr_by_year_rating <- group_by(loan, ProsperRating..Alpha., year, ProsperScore) %>%
                summarise(mean_apr= mean(BorrowerAPR),
                          mean_yield=mean(EstimatedEffectiveYield),
                          mean_amount=mean(LoanOriginalAmount),
                          mean_return=mean(EstimatedReturn),
                n = n()) %>%
                ungroup()



loan$ProsperScore <- factor(loan$ProsperScore,
                              levels = c(0,1,2,4,5,6,7,8,9,10))
apr_by_year_rating <- na.omit(apr_by_year_rating)
apr_by_year_rating$mean_apr <- round(apr_by_year_rating$mean_apr, digits =4) 
apr_by_year_rating$mean_return <- round(apr_by_year_rating$mean_return, digits =4) 


write.table(apr_by_year_rating, file="prosper_summary.tsv", quote=F, sep='\t', row.names=F)


```

```{r}
ggplot(aes(x=year, y= mean_apr, color=ProsperScore), data=apr_by_year_score) +
  geom_point() +
  scale_color_brewer(type='div') +
  theme_dark()
```
```{r}
ggplot(aes(x=year, y= mean_yield, color=ProsperRating..Alpha.), data=apr_by_year_rating) +
  geom_point() +
  scale_color_brewer(type='div') +
  theme_dark()
```

```{r}

years = c('2009', '2010', '2011', '2012', '2013', '2014')

for (year in years) {
 plot <- ggplot(aes(x=ProsperRating..Alpha., y= mean_apr, color=ProsperScore), data=apr_by_year_rating[apr_by_year_rating$year == year, ]) +
  geom_point() +
  scale_color_brewer(type='div') +
  theme_dark()
 
 print(plot)
}



```


```{r}

years = c('2009', '2010', '2011', '2012', '2013', '2014')

for (year in years) {
 plot <- ggplot(aes(x=ProsperRating..Alpha., y= mean_return, color=ProsperScore), data=apr_by_year_rating[apr_by_year_rating$year == year, ]) +
  geom_point() +
  scale_color_brewer(type='div') +
  theme_dark()
 
 print(plot)
}



```
```{r}
for (year in years) {
 plot <- ggplot(aes(x=ProsperRating..Alpha., y= mean_return, color=ProsperScore), data=apr_by_year_rating[apr_by_year_rating$year == year, ]) +
  geom_point() +
  scale_color_brewer(type='div') +
  theme_dark()
 
 print(plot)

}


```

```{r}

year_2009 <- loan[loan$year == '2009', ]


years = c('2009', '2010', '2011', '2012', '2013', '2014')

for (year in years) {
 plot <- ggplot(aes(x=ProsperRating..Alpha.), color=ProsperRating..Alpha.= data=loan[loan$year == year, ]) +
  geom_bar(aes(y = (..count..)/sum(..count..)))
 
 print(plot)
}

plot <- ggplot(aes(x=ProsperRating..Alpha.), color=ProsperRating..Alpha.= data=loan[loan$year == year, ]) +
  geom_bar(aes(y = (..count..)/sum(..count..)))

```
```{r}

by_year <- group_by(loan[], ProsperRating..Alpha., year) %>%
          summarise(count=n())


ggplot(aes(x=year, y=count, color=ProsperRating..Alpha.), data=by_year) +
  geom_line(aes(group=ProsperRating..Alpha.))

```

```{r}

years = c('2009', '2010', '2011', '2012', '2013', '2014')

for (year in years) {
 plot <- ggplot(aes(x=mean_apr, y= mean_return, color=ProsperRating..Alpha.), data=apr_by_year_rating[apr_by_year_rating$year == year, ]) +
  geom_point(aes()) +
  scale_color_brewer(type='div') +
  theme_dark()
 
 print(plot)
}



```
## Including Plots

You can also embed plots, for example:

```{r}
ggplot(aes(x=ProsperRating..Alpha., y= mean_return/mean_apr, color=ProsperRating..Alpha.), data=apr_by_year_rating) +
  geom_point() +
  scale_color_brewer(type='div') +
  theme_dark()

```




```{r}

year_2009 <- loan[loan$year == '2009', ]

ggplot(aes(x=ProsperScore, y=ProsperRating..numeric.), data=loan) +
  geom_point()

corr_test <- loan[, c('ProsperScore', 'ProsperRating..numeric.')]
corr_test <- na.omit(corr_test)

transform(corr_test, ProsperScore= as.double(ProsperScore))

typeof(score)

score <- corr_test$ProsperScore
rating <- corr_test$ProsperRating..numeric.



cor(score, rating)

```






```{r pressure, echo=FALSE}
str(loan)

#not sure why this logic is taking forever and 
start <- c(0)
end = c(10)
max = c(length(names(loan)))
while (end != max) {
  if (end > max) {
    end <- max
  }
  plot <- loan[, start:end]
  a <- ggplot(data = melt(plot), mapping = aes(x = value)) + 
    geom_histogram() +
    geom_density() + 
    facet_wrap(~variable, scales = 'free_x')
  print(a)
  start <- end
  end <- end + 10
}

summary(loan$CurrentCreditLines)
summary(loan$EmploymentStatusDuration)
summary(loan$PublicRecordsLast10Years)
summary(loan$StatedMonthlyIncome)
summary(loan$LoanOriginalAmount)
summary(loan$InvestmentFromFriendsCount)
summary(loan$Investors)
```
Notes about distributions

Borrow Rate is between .05 and .35
APR is slightly higher

looks like mostly 3 and 5 year terms 

despite different yields the estimate return is more tightly clustered which probably factors in that higher interest rates are more likely to default.

prosper score seems pretty evenly distributed with but with higher numbers in the middle of the spectrum

employment duration is skewed to the left. will have to drill down farther into it. 

way more credit lines than curret/ open than i though
summary(loan$OpenCreditLines)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   0.00    6.00    9.00    9.26   12.00   54.00    7604 
   
   
median is about $60000 a year then
summary(loan$StatedMonthlyIncome)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
      0    3200    4667    5608    6825 1750000 
      
spikes around multiples of 5
 summary(loan$LoanOriginalAmount)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   1000    4000    6500  
  8337   12000   35000 
  
Practically no loans are invested from a friends account
summary(loan$InvestmentFromFriendsCount)
  Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
0.00000  0.00000  0.00000  0.02346  0.00000 33.00000


loans from Investors is skewed left
summary(loan$Investors)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   1.00    2.00   44.00   80.48  115.00 1189.00 
```{r}
#this is not right i will have to group by 

library(dplyr)

summary(loan$BorrowerAPR)


apr_by_income <- group_by(loan, IncomeRange) %>%
                summarise(mean_apr= mean(BorrowerAPR),
                n = n())

apr_by_score <- group_by(loan, ProsperScore) %>%
                summarise(mean_apr= mean(BorrowerAPR),
                n = n())


apr_by_income <- subset(apr_by_income, !is.na(mean_apr))


ggplot(aes(x =IncomeRange, y=mean_apr), data = apr_by_income) + 
    geom_bar(stat = "identity") +
    scale_y_continuous(breaks = seq(0, .3, .05))


ggplot(aes(x =ProsperScore, y=mean_apr), data = apr_by_score) + 
    geom_bar(stat = "identity") +
    scale_y_continuous(breaks = seq(0, .3, .05))

ggplot(aes(x =ProsperRating..Alpha., y=mean_apr), data = apr_by_score) + 
    geom_bar(stat = "identity") +
    scale_y_continuous(breaks = seq(0, .3, .05))


    

apr_by_income$IncomeRange

is.ordered(apr_by_income$IncomeRange)


apr_by_income$IncomeRange <- factor(apr_by_income$IncomeRange,
                              levels = c('Not employed', '$0', '$1-24,999', '$25,000-49,999', '$50,000-74,999', '$75,000-99,999', '$100,000+'),
                              ordered = TRUE)
```



```{r, fig.width=12, fig.height=12}

summary(loan$LoanStatus)
```



First 
```{r, fig.width=12, fig.height=12}



loan_matrix <- loan[, c(5,6,8,9,10,11,12,13,14)]
loan_matrix <- na.omit(loan_matrix)


ggpairs(loan_matrix)





```

```{r}

ggplot(aes(x =EstimatedLoss, y=BorrowerAPR, color=LoanStatus), data = loan) + 
    geom_point()






```



```{r}

loan$ProsperScore <- factor(loan$ProsperScore,
                              levels = c(0,1,2,4,5,6,7,8,9,10))

ggplot(aes(x =EstimatedEffectiveYield, y=BorrowerAPR, color=IncomeRange), data = loan) + 
    geom_point()







```

```{r}

ggplot(aes(x=BorrowerAPR, y=EstimatedEffectiveYield, color=ProsperRating..numeric.), data = loan) + 
    geom_point(alpha = 1/10)






```
Key variables in the scorecard are:

Number of inquiries on the credit bureau
Number of delinquent accounts on the credit bureau
Credit card utilization on the credit bureau
Number of recently opened trades on the credit bureau
Debt to income ratio
Loan payment performance on prior Prosper loans
```{r, fig.width=12, fig.height=12}

loan$ProsperScore <- factor(loan$ProsperScore,
                              levels = c(0,1,2,4,5,6,7,8,9,10),
                            ordered = TRUE)

loan$Term <- factor(loan$Term, levels = c(12, 36, 60),
                    ordered = TRUE)

ggplot(aes(x =BorrowerAPR, y=EstimatedEffectiveYield, color=Term), data = loan) + 
    geom_point(aes(size=LoanOriginalAmount), alpha= 1/10)  +
    facet_wrap(~ProsperScore)

summary(loan$LoanOriginalAmount)


```




```{r}

loan$IncomeRange <- factor(loan$IncomeRange,
                              levels = c('Not employed', '$0', '$1-24,999', '$25,000-49,999', '$50,000-74,999', '$75,000-99,999', '$100,000+'),
                              ordered = TRUE)

ggplot(aes(x =EstimatedEffectiveYield, y=BorrowerAPR, color=IncomeRange), data = loan) + 
    geom_point() +
    scale_color_brewer(type ="seq")


```


```{r}
ggplot(aes(x =EmploymentStatusDuration, y=BorrowerAPR), data = loan) + 
    geom_point()


```


```{r}
ggplot(aes(x =ProsperRating..Alpha., y=BorrowerAPR, color=ProsperScore), data = loan) + 
    geom_point(alpha= 1/5) +
    scale_color_brewer(type='div')
    facet_wrap(~ProsperRating..Alpha.)




```


```{r}
ggplot(aes(x =ProsperScore, y=BorrowerAPR), data = loan) + 
    geom_boxplot()


```

Key variables in the scorecard are:

Number of inquiries on the credit bureau
Number of delinquent accounts on the credit bureau
Credit card utilization on the credit bureau
Number of recently opened trades on the credit bureau
Debt to income ratio
Loan payment performance on prior Prosper loans
```{r}
ggplot(aes(x =LoanOriginalAmount, y=BorrowerAPR), data = loan) + 
    geom_point()


```


Resources
https://www.orchardplatform.com/blog/alphabetic-ratings-and-numeric-scores-on-prosper/
https://www.prosper.com/about-us/wp-content/uploads/Performance_Update_January2017.pdf

http://dimplejs.org/advanced_examples_viewer.html?id=advanced_storyboard_control
https://github.com/d3/d3-scale-chromatic

http://colorbrewer2.org/#type=diverging&scheme=BrBG&n=7

https://discussions.udacity.com/t/custom-tooltip-legend-order/189042/6

https://techcrunch.com/2008/11/26/sec-outlines-its-reasoning-for-shutting-down-p2p-lender-prosper/